ohun: an R package for diagnosing and optimizing automated sound event detection

### 2023-02-14 manuscript/manuscript.Rmd
update figs and manuscript

Animal acoustic signals are widely used in a diversity of research areas. This is partly due to the relative ease with which sounds can be registered across a wide range of taxonomic groups and research settings. However, bioacoustics research can easily generate large amounts of data which might prove challenging to analyze in a timely manner. Many tools for the automated detection of sounds are currently available. Choosing the right approach for a specific task can be difficult as there are no available tools that can provide a common framework for evaluating detection performance. Here we present *ohun* an R package intended to facilitate the automated detection of sound events. *ohun* provides functions to diagnose and optimize detection routines as well as to compare the performance of different detection approaches. The package makes use of reference annotations containing the time position of target sounds in a training data set to evaluate the performance of detection routines, using common signal detection theory indices. This can be done with routine outputs imported from other software and from detection run within the package itself. The package also provides a set of functions to organize acoustic data sets in a format amenable for detection analyses. It also includes an implementation of two automatic detection methods commonly used in bioacoustic analysis: energy-based detection and template-based detection. We show how ohun can be used to automatically detect vocal signals with case studies of adult male zebra finch (*Taenopygia gutata*) songs and Spix's disc-winged bat (*Thyroptera tricolor*) ultrasonic social calls. We also include examples of how to evaluate the detection performance of ohun and external software. Finally, we provide some general suggestions to improve detection performance.

@@

The *ohun* package offers additional tools to simplify sound event detection. Detected sounds can be labeled as false or true positives with the function 'label_detection'. This allows users to explore the structure of false positives and figure out ways to exclude them. The function 'filter_detections' can be used to remove ambiguous sounds (i.e. those labeled as split or merged detections) keeping only those that maximize a specific criterium (i.e. the highest template correlation). Finally, note that several templates representing the range of variation in signal structure can be used to detect semi-stereotyped sounds when running template-based detection ('template_detection' function).

>>>>

Animal acoustic signals are widely used in diverse research areas due to the relative ease with which sounds can be registered across a wide range of taxonomic groups and research settings. However, bioacoustics research can quickly generate large data sets, which might prove challenging to analyze promptly. Although many tools are available for the automated detection of sounds, choosing the right approach can be difficult only a few tools provide a framework for evaluating detection performance. Here we present ohun, an R package intended to facilitate automated sound detection. ohun provides functions to diagnose and optimize detection routines and compare performance among different detection approaches. The package uses reference annotations containing the time position of target sounds in a training data set to evaluate detection routines performance using common signal detection theory indices. This can be done both with routine outputs imported from other software and detections run within the package. The package also provides functions to organize acoustic data sets in a format amenable to detection analyses. ohun also includes energy-based and template-based detection methods –two commonly used automatic approaches in bioacoustic research. We show how ohun automatically can be used to  detect vocal signals with case studies of adult male zebra finch (Taenopygia gutata) songs and Spix’s disc-winged bat (Thyroptera tricolor) ultrasonic social calls. We also include examples of how to evaluate the detection performance of ohun and external software. Finally, we provide some general suggestions to improve detection performance.

@@

The *ohun* package offers additional tools to simplify sound event detection. Detected sounds can be labeled as false or true positives with the function 'label_detection'. This allows users to explore the structure of false positives and figure out ways to exclude them. The function 'filter_detections' can be used to remove ambiguous sounds (i.e. those labeled as split or merged detections) keeping only those that maximize a specific criterion (i.e. the highest template correlation). Finally, note that several templates representing the range of variation in signal structure can be used to detect semi-stereotyped sounds when running template-based detection ('template_detection' function).
